A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-04-01 masters_thesis/si_paper/paper.tex
Removed some old files
Getting closer...

We ran experiments to compare our speculative parallelization of PSO to the
standard parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation for one particle, be it
speculative or non-speculative.  The speculative version actually performs two
iterations of PSO at each ``iteration,'' so we instead call each ``iteration''
a ``time step.''  As noted in the introduction, this is a reasonable
architecture to use if the function being optimized takes longer than around
100 milliseconds to run in our implementation.

Because our interest is in exploring the most efficient use of processors, we
must hold fixed the number of processors used at each time step in all of our
experiments, giving all approaches the same number of function evaluations.
Thus while we label our graphs with ``Time Steps,'' this is equivalent to the
number of function evaluations used by each approach (simply multiply by the
number of processors to transform time steps to function evaluations).  We use
``Time Steps'' to reflect wall-clock time, as our primary interest is using
extra processors in parallel to improve performance on functions with long
evaluation times.  A smaller swarm running serial PSO would likely produce
better results than we present in the same number of function evaluations, but
because the function evaluations were not done in parallel it would take more
wall-clock time.  When comparing the performance of parallelized algorithms on
long-running functions, we must compare by wall-clock time with the same
resources, not solely by the number of function evaluations.

Using the same number of processors for each approach requires that our
speculative parallelizations have a smaller swarm size than the standard
parallelization.  For each of the topologies we used, a particle has three
neighbors including itself.  As shown in Table~\ref{tab:evals}, this results in
$7$ speculative evaluations per particle.  With one evaluation needed for the
original, non-speculative particle, we have $8p$ evaluations for every two
iterations, where $p$ is the number of particles in the speculative swarm.  The
extra evaluations required in our speculative approach would instead be used to
evaluate particles in standard parallelizations, so we compare swarms of size
$p$ in speculative evaluation with swarms of size $8p$ in standard approaches.

\subsection{Running Time}

We do not report absolute time used by each algorithm, as specific time values
are heavily implementation-dependent.  Using time steps, where at every time
step each processor evaluates the function once, is an abstraction from
individual implementations that still conveys all of the information necessary
to compare algorithms.  In this paper we focus on the effects of algorithm
choices on parallel PSO performance.  As there are inexpensive functions with
complex landscapes and expensive functions with simple landscapes, the behavior
of PSO with respect to the function is the main issue.

For this purpose, benchmark functions are a useful and efficient tool for
understanding the effects of PSO with expensive objective functions even though
the benchmark functions are themselves inexpensive.  While benchmark functions
themselves take fractions of a second to evaluate and thus have no need of the
kind of parallelization we explore in this paper, they also have no need of
particle swarm optimization at all, as they can be solved analytically.  Yet
they are useful to us and to optimization researchers generally because they
stand as surrogates for the kinds of functions practitioners are actually
interested in, and they allow us to explore the behavior of optimization
algorithms on several different classes of functions in an easy and
standardized way.

While the discussion of exact running times of function evaluations and
communication overhead is appropriate, even essential, to a discussion of
specific parallel implementations, the concern of this paper is more
fundamental than implementation decisions.  One would not expect to see actual
running times in place of function evaluations in a paper describing an
innovation in the motion equations of PSO.  Similarly, actual running times in
a paper on algorithmic decisions in parallel PSO cloud the more important
issues of implementation-independent properties of the innovations described.

That being said, given the time required for the evaluation of the objective
function and the communication overhead per iteration for a specific
implementation, the actual running time can be recovered from all of our
results.  Simply multiply the number of time steps by the sum of the function
evaluation time and the overhead.  We wish to stress that the ``time steps'' we
report are proportional to wall-clock time, given the architecture we have
assumed.

@@

We also tested the parallelization methods on a real problem, that of fitting a
model to a large quantity of data.  We generated 10,000 data points from a
radial basis function network with 10 bases and some added Gaussian noise.  We
then fit a radial basis function network to the data using PSO.  This amounted
to a 30 dimensional function to optimize, with a function evaluation time on
the order of two seconds.

@@

The parallelization techniques we compare are the standard parallelization
(here labeled Standard simply for ease of reference), our original speculative
approach (labeled SEPSO), and the four relaxations of SEPSO discussed in
\sec{relax}.  In presenting our results, we call the approach developed in
\sec{pickbest} Pick Best.  The methods described in \sec{pruning} through
\sec{wrong} are called Pick Best Pruned and Social Promotion Pruned, and the
method in \sec{manyiters} we call Many Iterations.

@@

Because pruning only requires two speculative evaluations per particle (along
with evaluating the original particle), we can use swarms of size
$\frac{1}{3}p$ to compare to a swarm of size $p$ with the standard
parallelization, instead of $\frac{1}{8}p$ with other techniques.

@@

For each benchmark function we report results using the topology that is widely
considered best for that function, as reported in the literature
(e.g.,~\cite{bratton-2007-defining-a-standard-for-pso}).  In this paper we
limit ourselves to the Ring topology and the Complete topology, as is common
practice, along with the parallel approximation to the Complete topology
mentioned in \sec{pso}.  We also mentioned in \sec{related} that some related
work can be described as changes in topology, particularly that of having
subswarms of fully connected topologies that occasionally communicate with
each other.  This related work focused mainly on functions with very fast
evaluation times where such techniques drastically reduce interprocessor
communication.  With long function evaluations these topologies are not as
practical, as only one particle is evaluated by each processor at each
iteration.  We experimented with a subswarm topology and found that in every
instance except one it performed worse than either Complete or Ring, and thus
we leave it out of the results except in the one instance where it improved
performance.

Where the Complete topology would normally be preferred, we use a Random
topology in SEPSO and Pick Best, as Complete leads to an explosion in the
number of speculative evaluations (as noted, the other speculative techniques
can still use a Complete topology; we often report results on both topologies
for those methods).  If speculative evaluation were not being performed, it is
possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
distributed PSO, so it is still quite possible that Random would be used even
with standard parallelizations~\cite{mcnabb-2009-large-particle-swarms}.  But,
to be fair in our comparisons, we compare to the standard parallelization using
both the Random topology and the Complete topology (labeled PSO Random and PSO
Complete in our results).  Again, the amount of communication overhead is
heavily dependent on implementation details which are not the focus of this
paper.  A practitioner using a particular implementation can compare results
for whichever topology is most practical given the specific implementation.

@@

@@

@@

Rastrigin is a multi-modal function that is best optimized with a large,
Complete swarm.  It has been shown that with Rastrigin, the more particles
there are in the swarm, the lower function value it finds, up to at least 4000
particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
in local optima.  Because our speculative algorithms require significantly
smaller swarm sizes, we would expect to not perform very well on functions such
as Rastrigin.  Our experiments show our intuition to be correct.  In this
experiment we used 240 processors, so SEPSO and Pick Best each had 30
particles, and the Standard algorithms had 240 particles.  As expected, SEPSO
and Pick Best converge quickly to worse local optima than Standard does.
\fig{rastrigin} shows the results graphically, and \tabref{rastrigin-20} shows
results for all of the methods we tried.

@@

It is generally recommended to use the Ring topology when optimizing the
Griewank function, as Complete is prone to premature convergence on a local
optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  The global optimum
of Griewank has a value of 0.  When most trials reach the global optimum but a
few get stuck, the resultant ``average value'' graph has a flat line that is
misleading.  Thus instead we present plots showing the percent of trials that
have passed some threshold at each time step.  The threshold we chose for this
case was $10^{-6}$, as that value is below any local optima and the swarm
always successfully reaches 0 once is passes that point.

We ran 50 trials of each experiment with Griewank, so that the curves are more
smooth.  We show results in \fig{basic-griewank1} for swarms of size 30 and 240
using the Ring topology.  One can see in the figure that the SEPSO approach
finds the optimum faster than the Standard approach.  However, because the
swarm size is so small, SEPSO gets stuck a little less than half of the time.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank1}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step.  Recall that at each time step all processors evaluate the function
  once.  Instead of showing average function value, we show the percentage of
  runs that have found the global optimum by each iteration.  All algorithms
  use the Ring topology.}
  \label{fig:basic-griewank1}
\end{figure}

Because the speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{basic-griewank2} shows the results.
\fig{basic-griewank2} is very similar to \fig{basic-griewank1}, except that
speculative evaluation finds the global optimum essentially 100\% of the time.
Thus with a few more processors, our algorithm finds the optimum on average
twice as fast as the Standard approach without a significant loss of accuracy.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank2}
  \caption{Function Griewank with a swarm that uses 800 processors per time
  step.  All algorithms use the Ring topology.}
  \label{fig:basic-griewank2}
\end{figure}

In order to show the premature convergence properties of the various approaches
we have presented, we will show results first with swarms using 240 processors,
where these are more visible.  Then we will present an overall result for 800
processors, where using our techniques is generally more successful.

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank3}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step, comparing Pick Best with previous results.  All algorithms use the Ring
  topology.}
  \label{fig:griewank-pickbest}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank4}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step, comparing pruning with previous results.  All algorithms use the Ring
  topology.}
  \label{fig:griewank-pruned}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank5}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step, comparing speculating several iterations ahead with previous results.
  All algorithms use the Ring topology.}
  \label{fig:griewank-manyiters}
\end{figure}

Now that we have seen the effect that the various approaches have on speed of
convergence and success rates, we show a result for using 800 processors, where
most of the approaches have a 100\% success rate.  In this case, the best
algorithm to use is the one which finds the optimum the fastest.  As can be
seen in \fig{griewank-800}, with 800 processors the best choice is Pick Best,
which finds the optimum on average three times as fast as standard PSO.  With
just a few more processors, however, Many Iterations would likely have a 100\%
success rate as well, making it the preferred choice.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank6}
  \caption{Function Griewank with a swarm that uses 800 processors per time
  step.  All algorithms use the Ring topology.}
  \label{fig:griewank-800}
\end{figure}

Next we go to Sphere, the simplest of common benchmark functions.  The function
has no local optima (only a single global optimum) and is best optimized in
terms of function evaluations with a small swarm using a Complete topology.  We
expect our methods to be perfect for such functions, and our results show this
intuition to be correct.  We compared too many methods here to fit into one
graph, so we show just a few methods in \fig{sphere-20} and in
\tabref{sphere-20} we show a summary of the results for all methods we tested.

In \fig{sphere-20} we see that Many Iterations using a Complete topology
converges incredibly quickly on a very poor value.  We found this behavior to
be quite consistent across functions, so we rarely show results for Many
Iterations Complete.  However, Pick Best Pruned works very well with Complete
on this function.

Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
more by larger swarms than Sphere is.  Thus our speculative algorithms often
fail outperform Standard with 240 processors because the simple speculative
methods only have 30 particles.  As we will show later, when we use 800
processors at 50 dimensions, our methods perform much better.  However, the
pruned versions of our methods have 80 particles and thus are able to
outperform Standard even with only 240 processors.  Because the graph of
Schwefel 2.21 looks very similar to that of Sphere, we present all of our
results in \tabref{schwefel-20}.

@@

\begin{table}
  \caption{Summary of results for function Bohachevsky with 20 dimensions,
  measuring number of time steps to reach a value of 1e-06.}
  \label{tab:bohachevsky-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Standard Random&100\%&472.3&6.6\\
  \hline
  Standard Complete&100\%&238.2&11.2\\
  \hline
  SEPSO Random&75\%&261.6&9.2\\
  \hline
  Pick Best Random&100\%&211.8&9.2\\
  \hline
  Pick Best Pruned Random&100\%&189.7&3.4\\
  \hline
  Pick Best Pruned Complete&FIXME!&FIXME!&FIXME!\\
  \hline
  Social Promotion Pruned Random&100\%&260.3&6.5\\
  \hline
  Social Promotion Pruned Complete&FIXME!&FIXME!&FIXME!\\
  \hline
  \textbf{Many Iterations Random}&\textbf{94\%}&\textbf{118.2}&\textbf{2.6}\\
  \hline
  Many Iterations Complete&0\%&N/A&N/A\\
  \hline
  \end{tabular}
\end{table}

@@

Tables~\ref{tab:rastrigin-50} through~\ref{tab:bohachevsky-50} present
summaries of our results for each of the benchmark functions. As with the 20
dimensional variants, our speculative methods did not perform well on
Rastrigin, though for all other benchmarks Many Iterations significantly
outperformed Standard.  In each table, we bold the best method, meaning that it
has at least a 90\% success rate and its average time to completion is faster
than all other methods that have at least a 90\% success rate.  For example,
Many Iterations Random, when it is successful, achieves the required value 6
times faster than Standard Random.  However, because Many Iterations Random
only has a 25\% success rate while Standard Random has a 100\% success rate,
Standard Random is bolded.

We used a double-sided t-test to test for statistical significance in these
results, and all bolded values are statistically significant with a p-value of
less than $10^{-5}$.  In one instance two values are bolded because they are
statistically tied, while both are significantly better than all other methods.

@@

At 500 dimensions we found that the performance of constricted PSO on benchmark
functions becomes rather dismal.  In almost all cases, Standard fails to make
any significant progress.  Our brief search for an explanation seems to say
that the space is too large for particles to converge to the same point, so all
updates to the best position found come from a single particle wandering in the
space.  Every time that single particle finds a new best position, its velocity
contracts, so the particle is unable to make significant progress on its own.
In constricted PSO it is necessary to have a collection of particles exploring
a promising location to keep the particles' velocities from contracting
prematurely, and in 500 dimensions the space is too large to get the collection
of particles to the same location.

@@

For example, suppose a particle finds a new good position and updates its
values for $\pbest$ and $\nbest$, and that its velocity is such that it will
continue updating $\pbest$ and $\nbest$ if it continues in the same direction.
At that point, its velocity contracts, due to the motion equations in
constricted PSO (the social and cognitive components drop to zero and all that
is left is momentum, which is then constricted).  However, at that point we
speculate three iterations ahead along a path that is in roughly the same
direction the particle was just going, without constricting its velocity at
each of those iterations.  Thus the particle is able to make progress in that
direction instead of contracting its velocity at every iteration.

@@

For the model fitting problem that we experimented with, we used 144
processors.  We show results for three methods: Standard with a Random
topology, Standard with a Subswarm topology, and Many Iterations with a Random
topology.  Standard with a Subswarm topology had eight independent subswarms of
18 particles each, and the particles in each subswarm were connected with a
Random topology.  In the graphs we call this method Standard Subswarms.

\fig{rbf} shows our results for this function.  The function value reported is
the sum squared error of the model fit.  \fig{rbf-stuck} shows the percent of
runs that reached a value for sum squared error of 55,000 by each time step,
which we designate as successful.  Many Iterations took on average 126 time
steps to reach this value, while Standard Subswarms took on average 298.  The
p-value for a t-test on this statistic is less than $10^{-8}$.  With our
implementation of parallel PSO, each time step took on average 10.06 seconds;
1.83 seconds of that was function evaluation time and the rest was overhead.
With this particular implementation, then, using Many Iterations instead of
Standard led to an average time savings of 29 minutes, which is more than half
of the 50 minutes that Standard takes on average to reach a successful value.

@@

We have detailed how to perform speculative evaluation in PSO in several
different parallel frameworks.  Using this methodology, the behavior of the
original PSO algorithm can either be exactly reproduced, two iterations at a
time, or the behavior can be modified in order to improve performance.  While
exactly reproducing PSO often uses too many extra processors to be useful, when
we allow ourselves some freedom with the algorithm we see great improvements
over previous methods.  We have shown results that conclusively demonstrate the
superiority of our techniques for several functions over the standard practice
of adding particles to the swarm when extra processors are available.

@@

We have given five different possible approaches to speculative evaluation,
each of which has different properties.  These approaches perform differently
on different functions and at different swarm sizes, as would be expected by
the No Free Lunch Theorem~\citep{wolpert-1997-nfl-for-optimization}.  We have
given a brief evaluation of the premature convergence properties of these
methods on deceptive functions when a smaller number of processors is
available.  We have also shown evidence that when many processors are available
by far the best thing to do in most cases is to speculate as far ahead as the
extra processors allow.

Though our methods show great improvements on some functions, they do not work
for all functions.  As is commonly known, in PSO there is a trade-off between
exploration and exploitation.  Some functions need only minimal exploration,
and some never seem to have enough.  Increasing the swarm size is a natural way
to increase exploration in a parallel environment.  However, once ``enough''
exploration has been reached for any particular function, adding additional
particles adds only incremental benefits.  A better use of the additional
processors, as we have shown, is to perform some amount of speculative
evaluation.

>>>>

We ran experiments to compare our speculative parallelization of PSO to the
standard parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation for one particle, be it
speculative or non-speculative.  The exact evaluation time at which this
architecture becomes reasonable depends on the amount of communication overhead
in the parallel implementation and the number of particles in the swarm.  For
our implementation we found that time to be around 100 milliseconds for swarms
of 50 particles.  When the swarm size increases, the minimum evaluation time at
which this architecture should be used decreases.

\subsection{Function Evaluations vs. Time Steps}

Results in serial PSO are typically presented in terms of function evaluations.
This is a natural abstraction from implementation details that still allows a
comparison of the implementation-independent aspects of each algorithm.  The
number of function evaluations performed is assumed to be proportional to
wall-clock time, as all evaluations are performed in serial.  Only reporting
function evaluations could hide the fact that one algorithm requires more
overhead than another and thus actually takes more time to perform the same
number of function evaluations; however, function evaluations are still
considered the standard method of reporting, as evaluation times for functions
also vary greatly and could make the additional overhead negligible.

In parallel PSO on long-running functions, the natural way to present results
is in terms of iterations, not function evaluations.  This is because when all
function evaluations at each iteration are performed concurrently, iterations
are the direct equivalent of wall-clock time.  It is true that with a fixed
number of processors, function evaluations are also equivalent to wall-clock
time.  However, reporting function evaluations in a paper on parallel PSO can
be misleading if compared to a paper on serial PSO, as the actual wall-clock
times corresponding to the function evaluations are not directly comparable.
Thus we report iterations in our results instead of function evaluations.  But
because SEPSO actually performs two iterations of PSO at each ``iteration,''
and Social Promotion and Many Iterations make the idea of ``iterations of PSO''
somewhat nebulus, we instead call each ``iteration'' a ``time step.''

Just as serial PSO papers do not report actual running times of their specific
implementations, we do not report running times, favoring the abstraction of
time steps.  That being said, given the time required for the evaluation of the
objective function and the communication overhead per iteration for a specific
implementation, an actual running time can be recovered from all of our
results.  Simply multiply the number of time steps by the sum of the function
evaluation time and the overhead.  We wish to stress that the ``time steps'' we
report are proportional to wall-clock time, given the architecture we have
assumed.

@@

One might question the use of benchmark functions for evaluating our methods,
as they take on the order of 1 millisecond to evaluate instead of the 100s that
we said were required to justify our parallel architecture.  The purpose of
exploring the behavior of an algorithm on benchmark functions, whether in
serial or in parallel optimization research, is because benchmark functions are
believed to represent functions that one might encounter in the real world,
regardless of the amount of time those functions take to evaluate.  While
benchmark functions themselves take fractions of a second to evaluate and thus
have no need of the kind of parallelization we explore in this paper, they also
have no need of particle swarm optimization at all, as they can be solved
analytically.  Yet they are useful to us and to optimization researchers
generally because they stand as surrogates for the kinds of functions
practitioners are actually interested in, and they allow us to explore the
behavior of optimization algorithms on several different classes of functions
in an easy and standardized way.

To aid in convincing the reader that performance on benchmark functions roughly
corresponds to performance on real-world problems, we also tested our
parallelization methods on a typical research problem, that of fitting a model
to a large quantity of data.  We generated 10,000 data points from a radial
basis function network with 10 bases and some added Gaussian noise.  We then
fit a radial basis function network to the data using PSO.  This amounted to a
30 dimensional function to optimize, with a function evaluation time on the
order of two seconds.  We refer to this function as ``the model fitting
problem.''

@@

The parallelization techniques we compare are the standard parallelization
(here labeled Standard simply for ease of reference), our original speculative
approach (recall that we refer to this as SEPSO), and the four relaxations of
SEPSO discussed in \sec{relax}.  In presenting our results, we call the
approach developed in \sec{pickbest} Pick Best.  The methods described in
\sec{pruning} through \sec{wrong} are called Pick Best Pruned and Social
Promotion Pruned, and the method in \sec{manyiters} we call Many Iterations.

Using the same number of processors for each approach (and thus the same number
of function evaluations per time step) requires that our speculative
parallelizations have a smaller swarm size than the standard parallelization.
For the topologies we used with SEPSO and Pick Best, a particle has three
neighbors including itself.  As shown in Table~\ref{tab:evals}, this results in
$7$ speculative evaluations per particle.  With one evaluation needed for the
original, non-speculative particle, we have $8p$ evaluations for every two
iterations, where $p$ is the number of particles in the speculative swarm.  The
extra evaluations required in our speculative approach would instead be used to
evaluate particles in standard parallelizations, so we compare swarms of size
$p$ in speculative evaluation with swarms of size $8p$ in standard approaches.

@@

Because pruning only requires two speculative evaluations per particle (along
with evaluating the original particle), we can use swarms of size
$\frac{1}{3}p$ when pruning to compare to a swarm of size $p$ with the standard
parallelization, instead of $\frac{1}{8}p$ with other techniques.

@@

For each benchmark function we report results using the topology that is widely
considered best for that function, as reported in the literature
(e.g.,~\cite{bratton-2007-defining-a-standard-for-pso}).  In this paper we
limit ourselves to the Ring topology and the Complete topology, as is common
practice, along with the Random topology (the parallel approximation to the
Complete topology mentioned in \sec{pso}).  We also mentioned in \sec{related}
that some related work can be described as changes in topology, particularly
that of having subswarms of fully connected topologies that occasionally
communicate with each other.  This related work focused mainly on functions
with very fast evaluation times where such techniques drastically reduce
interprocessor communication.  With long function evaluations these topologies
are not as practical, as only one particle is evaluated by each processor at
each iteration.  We experimented with a subswarm topology and found that in
every instance except one it performed worse than either Complete or Ring, and
thus we leave it out of the results except in the one instance where it
improved performance.

Where the Complete topology would normally be preferred, we use a Random
topology in SEPSO and Pick Best, as Complete leads to an explosion in the
number of speculative evaluations (as noted, the other speculative techniques
can still use a Complete topology; we often report results on both topologies
for those methods).  If speculative evaluation were not being performed, it is
possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
distributed PSO, so it is still quite possible that Random would be used even
with standard parallelizations~\citep{mcnabb-2009-large-particle-swarms}.  But,
to be fair in our comparisons, we compare to the standard parallelization using
both the Random topology and the Complete topology (labeled PSO Random and PSO
Complete in our results).  Again, the amount of communication overhead is
heavily dependent on implementation details which are not the focus of this
paper.  A practitioner using a particular implementation can compare the
results given here for whichever topology is most practical given the specific
implementation.

@@

We frequently present tables summarizing our results.  In each table, we bold
the ``best'' method, meaning that it has at least a 90\% success rate and its
average time to completion is faster than all other methods that have at least
a 90\% success rate.  We used a double-sided t-test to test for statistical
significance in these results, and all bolded values are statistically
significant with a p-value of less than $10^{-5}$.  In some instances two
values are bolded because they are statistically tied, while both are
significantly better than all other methods.

@@

First we look at Sphere, the simplest of common benchmark functions.  The
function has a single global optimum and no other local optima.  Sphere is best
optimized in terms of function evaluations with a small swarm using a Complete
topology.  We expect our methods to be perfect for such functions, and our
results show this intuition to be correct.  For this comparison we used 240
processors, so the methods had swarms of size 30 (for SEPSO, Pick Best, and
Many Iterations), 80 (Pick Best Pruned and Social Promotion Pruned), and 240
(Standard).  We compared too many methods here to fit into one graph, so we
show just a few methods in \fig{sphere-20} and in \tabref{sphere-20} we show a
summary of the results for all methods we tested.

In \fig{sphere-20} we see that Many Iterations using a Complete topology
converges incredibly quickly on a very poor value.  We found this behavior to
be quite consistent across functions for this method, so we rarely show results
for Many Iterations Complete.  However, Pick Best Pruned works very well with
Complete on this function.

Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
more by larger swarms than Sphere is.  Thus our speculative algorithms often
fail to outperform Standard with 240 processors because the simple speculative
methods only have 30 particles.  As we will show later, when we use 800
processors at 50 dimensions, our methods perform much better.  However, the
pruned versions of our methods have 80 particles and thus are able to
outperform Standard even with only 240 processors.  Because the graph of
Schwefel 2.21 looks very similar to that of Sphere, we simply present all of
our results in \tabref{schwefel-20}.

@@

Rastrigin is a multi-modal function that is best optimized with a large,
Complete swarm.  It has been shown that with Rastrigin, the more particles
there are in the swarm, the lower function value it finds, up to at least 4000
particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
in local optima and converge to poorer values.  Because our speculative
algorithms require significantly smaller swarm sizes, we would expect to not
perform very well on functions such as Rastrigin.  Our experiments show our
intuition to be correct.  In this experiment we used 240 processors, so SEPSO
and Pick Best each had 30 particles, and the Standard algorithms had 240
particles.  As expected, SEPSO and Pick Best converge quickly to worse local
optima than Standard does.  \fig{rastrigin} shows the results graphically, and
\tabref{rastrigin-20} shows results for all of the methods we tried.

@@

It is generally recommended to use the Ring topology when optimizing the
Griewank function, as Complete is prone to premature convergence on a local
optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  The global optimum
of Griewank has a value of 0.  When most trials reach the global optimum but a
few get stuck, the resultant ``average value'' graph has a flat line that is
misleading.  Thus instead we present plots showing the percent of trials that
have passed some threshold at each time step, as is common practice with these
functions~\citep{mendes-2004-population-topologies-in-pso}.  The threshold we
chose for this case was $10^{-6}$, as that value is below any local optima and
the swarm always successfully reaches 0 once is passes that point.

We ran 50 trials of each experiment with Griewank, so that the curves are more
smooth.  We show results in \fig{basic-griewank1} for swarms of size 100 and
800 using the Ring topology.  One can see in the figure that SEPSO reaches the
global optimum on average close to twice as fast as Standard, while Pick Best
is close to three times as fast and Many Iterations is even faster when it is
successful, though it is only successful 84\% of the time.  \tabref{griewank}
shows all of our results in tabular form.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank6}
  \caption{Function Griewank with a swarm that uses 800 processors per time
  step.  Recall that at each time step all processors evaluate the function
  once.  Instead of showing average function value, we show the percentage of
  runs that have found the global optimum by each iteration.  All algorithms
  use the Ring topology.}
  \label{fig:basic-griewank1}
\end{figure}

\begin{table}
  \caption{Summary of results for function Griewank with 20 dimensions,
  measuring number of time steps to reach a value of 1e-06}
  \label{tab:griewank-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Standard Ring&100\%&762.2&19.4\\
  \hline
  SEPSO Ring&100\%&426.0&81.0\\
  \hline
  \textbf{Pick Best Ring}&\textbf{100\%}&\textbf{272.0}&\textbf{17.7}\\
  \hline
  \textbf{Pick Best Pruned Ring}&\textbf{100\%}&\textbf{282.9}&\textbf{10.7}\\
  \hline
  Social Promotion Pruned Ring&100\%&482.0&16.6\\
  \hline
  Many Iterations Ring&84\%&183.2&49.4\\
  \hline
  \end{tabular}
\end{table}

We pause here to show some interesting characteristics of the speculative
techniques we have proposed.  With 800 processors our methods perform very well
on this function.  With 240 processors, however, the results are much more
mixed.  Because 240 processors is near the point at which speculative
evaluation becomes useful, it is enlightening to see the behavior of the
various algorithms at this point.

We show results in \fig{basic-griewank2} for swarms of size 30 and 240 using
the Ring topology.  One can see in the figure that when SEPSO is successful, it
finds the optimum much faster than Standard.  However, because the swarm size
is so small, SEPSO gets stuck a little less than half of the time.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank1}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step.  All algorithms use the Ring topology.}
  \label{fig:basic-griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank3}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step, comparing Pick Best with results from \fig{basic-griewank2}.  All
  algorithms use the Ring topology.}
  \label{fig:griewank-pickbest}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank4}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step, comparing pruning with the best results from \fig{griewank-pickbest}.
  All algorithms use the Ring topology.}
  \label{fig:griewank-pruned}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank5}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step, comparing speculating several iterations ahead with selections from
  \fig{griewank-pruned}.  All algorithms use the Ring topology.}
  \label{fig:griewank-manyiters}
\end{figure}

@@

\begin{table}
  \caption{Summary of results for function Bohachevsky with 20 dimensions,
  measuring number of time steps to reach a value of 1e-06.}
  \label{tab:bohachevsky-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Standard Random&100\%&472.3&6.6\\
  \hline
  Standard Complete&100\%&238.2&11.2\\
  \hline
  SEPSO Random&75\%&261.6&9.2\\
  \hline
  Pick Best Random&100\%&211.8&9.2\\
  \hline
  Pick Best Pruned Random&100\%&189.7&3.4\\
  \hline
  Pick Best Pruned Complete&25\%&105.4&3.9\\
  \hline
  Social Promotion Pruned Random&100\%&260.3&6.5\\
  \hline
  Social Promotion Pruned Complete&90\%&197.9&12.2\\
  \hline
  \textbf{Many Iterations Random}&\textbf{94\%}&\textbf{118.2}&\textbf{2.6}\\
  \hline
  Many Iterations Complete&0\%&N/A&N/A\\
  \hline
  \end{tabular}
\end{table}

@@

Tables~\ref{tab:rastrigin-50} through~\ref{tab:bohachevsky-50} present
summaries of our results for each of the benchmark functions. As with the 20
dimensional variants, our speculative methods did not perform well on
Rastrigin, though for all other benchmarks Many Iterations significantly
outperformed Standard.

@@

At 500 dimensions we found that the performance of constricted PSO on benchmark
functions becomes rather dismal.  In almost all cases, Standard fails to make
any significant progress.  Our search for an explanation seems to say that the
space is too large for particles to converge to the same point, so all updates
to the best position found come from a single particle wandering in the space.
Every time that single particle finds a new best position, its velocity
contracts, so the particle is unable to make significant progress on its own.
In constricted PSO it is necessary to have a collection of particles exploring
a promising location to keep the particles' velocities from contracting
prematurely, and in 500 dimensions the space is too large to get the collection
of particles to the same location.

@@

@@

For the model fitting problem we used 144 processors.  We show results for
three methods: Standard with a Random topology, Standard with a Subswarm
topology, and Many Iterations with a Random topology.  Standard with a Subswarm
topology had eight independent subswarms of 18 particles each, and the
particles in each subswarm were connected with a Random topology.  In the
graphs we call this method Standard Subswarms.

\fig{rbf} shows our results for this function.  The function value reported is
the sum squared error of the model fit.  \fig{rbf-stuck} shows the percent of
runs that reached a value for sum squared error of 55,000 by each time step,
which we designate as successful.  Many Iterations took on average 126 time
steps to reach this value, while Standard Subswarms took on average 298.  The
p-value for a t-test on this statistic is less than $10^{-8}$.  With our
implementation of parallel PSO, each time step took on average 10.06 seconds;
1.83 seconds of that was function evaluation time and the rest was overhead.
With this particular implementation, then, Many Iterations takes on average 21
minutes to reach a successful value, while Standard takes 50 minutes.

@@

We have detailed how to perform speculative evaluation in PSO in several
different parallel architectures.  Using this methodology, the behavior of the
original PSO algorithm can either be exactly reproduced, two iterations at a
time, or the behavior can be modified in order to improve performance.  While
exactly reproducing PSO sometimes uses too many extra processors to be useful,
when we allow ourselves some freedom with the algorithm we see great
improvements over previous methods.  We have shown results that conclusively
demonstrate the superiority of our techniques for several functions over the
standard practice of adding particles to the swarm when extra processors are
available.

@@

We have given five different possible approaches to speculative evaluation,
each of which has different properties.  These approaches perform differently
on different functions and at different swarm sizes, as would be expected by
the No Free Lunch Theorem for
Optimization~\citep{wolpert-1997-nfl-for-optimization}.  We have given a brief
evaluation of the premature convergence properties of these methods on
deceptive functions when a smaller number of processors is available.  We have
also shown evidence that when many processors are available by far the best
thing to do in most cases is to speculate as far ahead as the extra processors
allow.

Though our methods show great improvements on some functions, they do not work
for all functions.  As is commonly known, in PSO there is a trade-off between
exploration and exploitation.  Some functions need only minimal exploration,
and some never seem to have enough.  Increasing the swarm size is a natural way
to increase exploration in a parallel environment.  However, once ``enough''
exploration has been reached for any particular function, adding additional
particles adds only incremental benefits.  At this point, a better use of the
additional processors, as we have shown, is to perform some amount of
speculative evaluation.
